In-Band Synchronization for\\Distributed SDN Control Planes\thanks{Supported by the European Institute of Innovation \& Technology (EIT) project 13153 (Software Defined Networking) and the FP7 EU project UNIFY.}

### 2015-11-20 ccr-camera/inband-sync.tex
camera ready proposal

\category{C.2.2}{Computer-Communication Networks}{Network Protocols}
\category{C.2.3}{Network Operations}{Network Management}
%\category{G.1.6}{Optimization}{Linear Programming}

\terms{Networks}

\keywords{Software-Defined Networking, Distributed Control Planes, In-band Mechanisms}

@@

@@

As a case study, and to demostrate the feasibility
of our conceptual approach, we 
consider the standard \emph{OpenFlow}
 protocol (version 1.4~\cite{openflow}). 
%\stefan{liron can we say that
% for example novikit supports all these features?}\liron{No. Novikit support OF 1.3 and therefore lacks support for bundles.}
% In particular, 
%PK ``key-value store'' come out of the blue...
We show that OpenFlow's match-action paradigm can be instrumented
to realize 
%simple key-value stores in the configuration space.
%
%\petr{Did not get the flow here: what is ``our abstraction'' here?
a simple transactional abstraction that allows 
the controllers to solve several important synchronization problems.
For example, we describe an OpenFlow implementation of the fundamental
\emph{compare-and-set} (CAS) primitive.
The atomic CAS operation is known to provide an infinite consensus number,
and can, for instance, be used to 
implement a generic consistent and fault-tolerant replicated 
%[[PK why control-plane here: does not fit the Chubby et al. mentioned later.
%control-plane 
service~\cite{Her91} 
%in a consistent and fault-tolerant way: 
a mainstream building block in
modern distributed systems (Chubby, Google Spanner, Amazon Webservices, etc.).
We discuss a simple CAS-based concurrent policy-update mechanism,
and present a generic template to implement concurrent policy
compositions of a user-specific update functions.
Our synchronization abstraction also provides
the missing link for the read-modify-write object
postulated in CPC and STN~\cite{infocom15}. %\petr{We do not seem to discuss the latter?}

@@

\section{Motivation and Example}\label{sec:motivation}

A transactional distributed system offers all-or-nothing semantics, 
and aborts transactions in case of conflicts. 
For instance, in the context of SDNs, a transaction trying to add a forwarding rule whose
domain overlaps with an existing rule of the same priority
can constitute a conflict. 
However, as we will argue, 
there exist many applications with less obvious conflicts, which go 
beyond simple range overlaps. 
Moreover, sometimes it is desirable to react to conflicts in smarter ways than
by simply aborting an operation, e.g., by supporting \emph{conditional modifications}.

@@

\vspace{1mm}
\noindent\textbf{Configuration space.}
%
We distinguish two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules),
and the \emph{meta-configuration} that contains information used by the
controllers to synchronize their actions.
Flow entries in the meta-configuration can be referred to as shared
memory locations, each provided with a distinct \emph{address}.

@@

As a case study, in this section, we show how to implement
our transactional approach using today's standard SDN protocol,
 OpenFlow.

\subsection{Background}

The configuration of an OpenFlow switch is a set of
\emph{rules}.
A rule is essentially a \emph{match-action} pair:
the match part of a rule identifies the space of packet headers that are
subject to the rule (e.g., all packets to a specific destination) and
the action part identifies how the switch should process the matched
packets (e.g., forward them to a specific port).
More precisely, the match part of a rule is
a ternary pattern over packet \emph{header} fields.
The pattern is represented as a \emph{value} and a \emph{mask}.
%\liron{we need to discuss about this, I am not sure this description is accurate.}
In the mask, certain header fields, e.g., TCP or UDP port numbers or destination and source
addresses, can be \emph{wildcarded}, stipulating that their content does
not affect the match.
Certain fields, such as \emph{ipv4\_src}, \emph{ipv4\_dst}, \emph{metadata}, can be arbitrarily
bitmasked.
In our implementations, we are going to use bitmasking of the \emph{metadata}
field for performing conditional operations based on  its content.
%[[PK put the example back in response of review 2
For example, \texttt{ipv4\_src=10.1.14.0/ff.ff.ff.f0} matches all packets
with source IP addresses in the range \texttt{10.1.14.0}--\texttt{10.1.14.15}.
%]]

@@

\subsection{Implementation}

At first sight, it may seem that bundling \emph{FlowMod} commands with  the $\checko$
flag already provides a useful synchronization mechanism.
Indeed, its ``mini-transaction'' nature allows a controller
to install multiple flow entries in an atomic way.
However, as is,
bundling has important limitations. %, as we will argue in the following.

@@

In order to check if an identifier $x$ is claimed
(Algorithm~\ref{alg:check}), a controller $c$ tries to add
a ``tester'' entry, where the match value field is 
$(c\concat*^{16}\concat x)$  and the flag \texttt{\checko} is set.
Note that the controller identifier is shifted $16$ bits to the left in
the match field, which allows us to ensures that the entry is uniquely
identified, while not conflicting with other entries.  
This attempt inflicts a failure in case another entry with claimed
value $x$ exists. 
If the check succeeds, i.e.,  no entry with claimed value $x$ is
present, the tester entry is deleted.   
%[[PK unnecessary?
%In any case since the check entry contains the controller id, it does not conflict with other check entries and there is no negative impact of delaying the deletion.
%]]
%Considering $m_1$ and $m_2$ from the example above,
%with respective actions $a_1=1$ and $a_2=2$,  they overlap
%with different actions iff $x_1=x_2$. Therefore, if $m_1$ is used in the \claimcheck after $m_2$ was used in the claim operation,
%a conflict is detected iff $x_1=x_2$, as desired.

@@

Let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (\pid)} .
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The CAS operation ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
Note that, since all policy modifications are executed within a
bundle providing controller-atomicity, the \textbf{read-state} command
in Line~\ref{line:readstate} appears atomic.  
Algorithm~\ref{alg:simple-update} can therefore also be used
to render a na\"ive, single-controller load-balancing algorithm
consistent, without any modifications in the update function.

@@

Algorithm~\ref{alg:simple-update} is correct assuming that policy
identifiers may grow without bound. 
In practice, we can use a module scheme on the 32-bit field of a flow entry
used for storing the policy identifier, which, in principle,
may lead to inconsistencies when the relative speeds of controllers is
very high.   
Algorithm~\ref{alg:update} adresses this issue 
by using our id-claimer abstraction.
%to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and 
claims it, which prevents another controller from using the id 
for a different policy
(Lines~\ref{update:read-state1}-\ref{update:read-state2}). 
Then the controller selects any \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}). %\liron{I have replaced "claims the first" with "chooses an" following Petr email}
Finally, within an atomic transaction (\textbf{\execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then tries to update the
policy.
%

@@

\vspace{1mm}
\noindent\textbf{A note on efficiency.}
%
We observe that our implementations of synchronization primitives
(Algorithms~\ref{alg:claim}-\ref{alg:compare}) 
%is \emph{switch-space efficient}
%in the sense that 
require only a constant number of flow entries per memory address or claim.
The number of used addresses is constant for  
Algorithm~\ref{alg:simple-update} and linear in the number of
controllers for Algorithm~\ref{alg:update}. 
%[[PK not constant: proportional to the number of failures. But this
%is kind of obvious.
%Also, in case of controller failures, at most a constant amount of space is lost.
%]]

@@

{\scriptsize
\begin{algorithm}[t]
    \caption{Advanced policy update}
    \label{alg:update}
    \begin{algorithmic}[1]
        \Require policy update function \textsc{\ufunc}, policy id address $\paddr$, id space $C$.
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
		 	\State $\pid,claims,policy\gets$ \textbf{read-state} \label{update:read-state1} %switch configuration
 			\State $\textbf{execute}\{claim(\pid)\}$
 			\State $\pid 2,\textit{claims},\textit{policy}\gets$ \textbf{read-state} \label{update:read-state2} %switch configuration
 			\If {$\pid\neq \pid 2$}
	 			\State $\textbf{execute} \{\textit{unclaim}(\pid)\}$
 				\State {\bf continue} (\textbf{restart} loop)
 			\EndIf
 			\State $\textit{my\_id}\gets$ choose a number from $C\setminus claims$
 			\State $\textit{update\_cmds}\gets \textsc{\ufunc}(policy)$ \label{update:compute}
 			\startTxn
 				\State $\claimcheck(my\_id)$,
	 			\State $CAS(\paddr, \pid,my\_id)$,
	 			\State $\textit{update\_cmds}$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
	 		\State $\textbf{execute} \{\textit{unclaim}(\pid)\}$
     		%\State $res \gets $ commit transaction
     	\Until{$res=\ack$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return $\textit{res}$

@@

To investigate the feasibility of our approach,
we implemented a proof-of-concept of our algorithms
as a python library which wraps ovs-ofctl, 
a well-known command line tool for monitoring and administering OpenFlow switches. 
The code is available online at
\url{https://github.com/lironsc/of-sync-lib}. 
%\liron{perhaps we can put it as reference? (putting the link in bib)}

@@

For instance, we experimented with a setting where three processes
(controllers), concurrently issuing $1000$ concurrent policy updates
each, where every update replaces a single rule with another one.
%[[PK unclear
%When verifying the transactions, 
%]]
We could confirm that 
despite the concurrency, 100\% of the updates 
eventually completed in committing their transactions, possibly after
several aborts caused by conflicts. 
%Of course, due to the concurrency, some transactions also needed to abort
%because of conflicts. 
%\petr{Reads strange: 100\% successfullt complete and some transactions
 % abort...}

@@

We believe that our centralized in-band
synchronization approach is natural and attractive, as it avoids the
complexities and limitations of distributed out-of-band agreement protocols, which may 
introduce additional points of potential failures.
Moreover, the switch is the element implementing the controller policies anyway.
The mechanisms presented in this paper are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{infocom15,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions~\cite{pyretic}.
We see our paper as a first step, and more
work is required to investigate and compare useful
in-band synchronization abstractions, also in terms of performance.
While our framework can be used as a synchronization library, we 
also work on integrating it into an existing SDN controller, e.g., Ryu.

>>>>

%\category{C.2.2}{Computer-Communication Networks}{Network Protocols}
\category{C.2.3}{Network Operations}{Network Management}
%\category{G.1.6}{Optimization}{Linear Programming}

%\terms{Networks}

%\keywords{Software-Defined Networking, Distributed Control Planes, In-band Mechanisms}

@@

\paragraph{Contributions}

@@

As a case study, and to demostrate the feasibility
of our conceptual approach, we 
consider the standard \emph{OpenFlow}
 protocol (version 1.4~\cite{openflow}). 
%\stefan{liron can we say that
% for example novikit supports all these features?}\liron{No. Novikit support OF 1.3 and therefore lacks support for bundles.}
% In particular, 
%PK ``key-value store'' come out of the blue...
We show that OpenFlow's match-action paradigm can be instrumented
to realize 
%simple key-value stores in the configuration space.
%
%\petr{Did not get the flow here: what is ``our abstraction'' here?
a simple transactional abstraction that allows 
the controllers to solve several important synchronization problems.
For example, we describe an OpenFlow implementation of the fundamental
\emph{compare-and-set} (CAS) primitive.
The atomic CAS operation is known to provide an infinite consensus number,
and can, for instance, be used to 
implement a generic consistent and fault-tolerant replicated 
%[[PK why control-plane here: does not fit the Chubby et al. mentioned later.
%control-plane 
service~\cite{Her91}, 
%in a consistent and fault-tolerant way: 
a mainstream building block in
modern distributed systems (Chubby, Google Spanner, Amazon Webservices, etc.).
We discuss a simple CAS-based concurrent policy-update mechanism,
and present a generic template to implement concurrent policy
compositions of a user-specific update functions.
Our synchronization abstraction also provides
the missing link for the read-modify-write object
postulated in STN~\cite{infocom15}. %\petr{We do not seem to discuss the latter?}

@@

\paragraph{Motivation and Example}
A transactional distributed system offers all-or-nothing semantics, 
and aborts transactions in case of conflicts. 
For instance, in the context of SDNs, a transaction trying to add a forwarding rule whose
domain overlaps with an existing rule of the same priority
can constitute a conflict. 
However, as we will argue, 
there exist many applications with less obvious conflicts, which go 
beyond simple range overlaps. 
Moreover, sometimes it is desirable to react to conflicts in smarter ways than
by simply aborting an operation, e.g., by supporting \emph{conditional modifications}.

@@

\vspace{1mm}
\noindent\textbf{Configuration space.}
%
We distinguish between two parts of the configuration space of a switch:
the ``normal'' configuration which determines the network policy and
is used to process the data-plane traffic (e.g., forwarding rules),
and the \emph{meta-configuration} that contains information used by the
controllers to synchronize their actions.
Flow entries in the meta-configuration can be referred to as shared
memory locations, each provided with a distinct \emph{address}.

@@

Before discussing applications in more detail, 
as a case study and in order to introduce some notation, 
we in this section show how to implement
our transactional approach using today's standard SDN protocol,
 OpenFlow  (version 1.4~\cite{openflow}).\footnote{We note
however that hardware OpenFlow switches do not always 
follow the spec as they should~\cite{know}, and will report on a
prototype implementation later in this paper.}

\paragraph{Background}
The configuration of an OpenFlow switch is a set of
\emph{rules}.
A rule is essentially a \emph{match-action} pair:
the match part of a rule identifies the space of packet headers that are
subject to the rule (e.g., all packets to a specific destination) and
the action part identifies how the switch should process the matched
packets (e.g., forward them to a specific port).
More precisely, the match part of a rule is
a ternary pattern over packet \emph{header} fields.
The pattern is represented as a \emph{value} and a \emph{mask}.
%\liron{we need to discuss about this, I am not sure this description is accurate.}
In the mask, certain header fields, e.g., TCP or UDP port numbers or destination and source
addresses, can be \emph{wildcarded}, stipulating that their content does
not affect the match.
Certain fields, such as \emph{ipv4\_src}, \emph{ipv4\_dst}, \emph{metadata}, can be arbitrarily
bitmasked.
In our implementations, we are going to use bitmasking of the \emph{metadata}
field for performing conditional operations based on  its content.
%[[PK put the example back in response of review 2
For example, \texttt{ipv4\_src=10.1.14.0/ff.ff.ff.f0} matches all packets
with source IP addresses in the range \texttt{10.1.14.0}--\texttt{10.1.14.15}.
%]]

@@

\paragraph{Implementation} 
At first sight, it may seem that bundling \emph{FlowMod} commands with  the $\checko$
flag already provides a useful synchronization mechanism.
Indeed, its ``mini-transaction'' nature allows a controller
to install multiple flow entries in an atomic way.
However, as is,
bundling has important limitations. %, as we will argue in the following.

@@

In order to check if an identifier $x$ is claimed
(Algorithm~\ref{alg:check}), a controller $c$ tries to add
a ``tester'' entry, where the match value field is 
$(c\concat*^{16}\concat x)$  and the flag \texttt{\checko} is set.
Note that the controller identifier is shifted $16$ bits to the left in
the match field, which allows us to ensure that the entry is uniquely
identified, while not conflicting with other entries.  
This attempt inflicts a failure in case another entry with claimed
value $x$ exists. 
If the check succeeds, i.e.,  no entry with claimed value $x$ is
present, the tester entry is deleted.   
%[[PK unnecessary?
%In any case since the check entry contains the controller id, it does not conflict with other check entries and there is no negative impact of delaying the deletion.
%]]
%Considering $m_1$ and $m_2$ from the example above,
%with respective actions $a_1=1$ and $a_2=2$,  they overlap
%with different actions iff $x_1=x_2$. Therefore, if $m_1$ is used in the \claimcheck after $m_2$ was used in the claim operation,
%a conflict is detected iff $x_1=x_2$, as desired.

@@

Let us consider a simple policy update protocol that
allows a set of  controllers to concurrently update switch policies, i.e., sets of
%[[PK policies have been defined in Background]]
effective flow entries, under the
condition that these rules can be \emph{composed} with the currently installed
policy.
Algorithm~\ref{alg:simple-update} uses the bundle feature to update
the switch configuration, where the configuration contains both the
policy and a meta-configuration memory address $\paddr$ that holds the \emph{policy identifier (\pid)}.
In the algorithm, the controller first reads the currently installed
policy together with its {\pid},  applies the
$\textsc{\ufunc}()$ function to the current policy which results in a
set of update FlowMod commands and then tries to apply them
atomically together with a CAS operation on $\paddr$ replacing $\pid$
with $\pid+1$.
The CAS operation ensures that if, concurrently, a new policy (with a different identifier) has been installed, the update
fails and takes no effect.
Note that, since all policy modifications are executed within a
bundle providing controller-atomicity, the \textbf{read-state} command
in Line~\ref{line:readstate} appears atomic.  
Algorithm~\ref{alg:simple-update} can therefore also be used
to render a na\"ive, single-controller load-balancing algorithm
consistent, without any modifications in the update function.

@@

Algorithm~\ref{alg:simple-update} is correct assuming that policy
identifiers may grow without bound. 
In practice, we can use a modulo scheme on the 32-bit field of a flow entry
used for storing the policy identifiers, which, in principle,
may lead to inconsistencies if controllers proceed in different speeds.
Algorithm~\ref{alg:update} adresses this issue 
by using our id-claimer abstraction.
%to bound the number of used policy identifiers.
Here, the controller first reads the current policy id and 
claims it, which prevents another controller from using the id 
for a different policy
(Lines~\ref{update:read-state1}-\ref{update:read-state2}). 
Then the controller selects any \emph{unused} id and computes the update commands to be executed (Line~\ref{update:compute}). %\liron{I have replaced "claims the first" with "chooses an" following Petr email}
Finally, within an atomic transaction (\textbf{\execatomic}), it verifies whether the chosen policy id has not
been claimed and that
the policy has not been concurrently modified, and then tries to update the
policy.
%

@@

\vspace{1mm}
\noindent\textbf{A note on efficiency.}
%
Our implementations of synchronization primitives
(Algorithms~\ref{alg:claim}-\ref{alg:compare}) 
%is \emph{switch-space efficient}
%in the sense that 
require only a constant number of flow entries per memory address or claim.
The number of used addresses is constant for  
Algorithm~\ref{alg:simple-update} and linear in the number of
controllers for Algorithm~\ref{alg:update}. 
%[[PK not constant: proportional to the number of failures. But this
%is kind of obvious.
%Also, in case of controller failures, at most a constant amount of space is lost.
%]]

@@

{\scriptsize
\begin{algorithm}[t]
    \caption{Advanced policy update}
    \label{alg:update}
    \begin{algorithmic}[1]
        \Require policy update function \textsc{\ufunc}, policy id address $\paddr$, id space $C$.
    \Ensure installed policy is consistent with previous one
 		%\Do
 		\Repeat
		 	\State $\pid,claims,policy\gets$ \textbf{read-state} \label{update:read-state1} %switch configuration
 			\State $\textbf{execute}\{claim(\pid)\}$
 			\State $\pid 2,\textit{claims},\textit{policy}\gets$ \textbf{read-state} \label{update:read-state2} %switch configuration
 			\If {$\pid\neq \pid 2$}
	 			\State $\textbf{execute} \{\textit{unclaim}(\pid)\}$
 				\State {\bf continue} (\textbf{restart} loop)
 			\EndIf
 			\State $\textit{my\_id}\gets$ choose a number from $\{C\setminus claims\}$
 			\State $\textit{update\_cmds}\gets \textsc{\ufunc}(policy)$ \label{update:compute}
 			\startTxn
 				\State $\claimcheck(my\_id)$,
	 			\State $CAS(\paddr, \pid,my\_id)$,
	 			\State $\textit{update\_cmds}$ %conditioned on $metadata=my\_a$
	 			%\State update policy selector to $my\_a$
 			\endTxn
	 		\State $\textbf{execute} \{\textit{unclaim}(\pid)\}$
     		%\State $res \gets $ commit transaction
     	\Until{$res=\ack$}
     	%\doWhile{$res=False$}
%   \vspace{3mm}
			\Return $\textit{res}$

@@

To investigate the feasibility of our approach,
we implemented a proof-of-concept of our algorithms
as a python library which wraps ovs-ofctl, 
a well-known command line tool for monitoring and administering OpenFlow switches. 
The code is made available online\footnote{See \url{https://github.com/lironsc/of-sync-lib}.}. 
%\liron{perhaps we can put it as reference? (putting the link in bib)}

@@

For instance, we experimented with a setting where three processes
(controllers), concurrently issue $1000$ concurrent policy updates
each, where every update replaces a single rule with another one.
%[[PK unclear
%When verifying the transactions, 
%]]
We could confirm that 
despite the concurrency, 100\% of the updates 
eventually completed in committing their transactions, possibly after
several aborts caused by conflicts. 
%Of course, due to the concurrency, some transactions also needed to abort
%because of conflicts. 
%\petr{Reads strange: 100\% successfullt complete and some transactions
 % abort...}

@@

We believe that our centralized in-band
synchronization approach is natural and attractive, as it avoids the
complexities and limitations of distributed out-of-band agreement protocols, which may 
introduce additional points of potential failures.
Moreover, the switch is the element implementing the controller policies anyway.
The mechanisms presented in this paper are simple and can be implemented using the \emph{standard OpenFlow}
protocol: they do not require any protocol/hardware extensions as postulated in recent literature~\cite{infocom15,netpaxos}.
Our work may hence also contribute to the ongoing discussion of what can be implemented
in-band in today's OpenFlow protocol~\cite{reclaim},
as well as to useful high-level concurrency objects and
language abstractions~\cite{pyretic}.
We see our paper as a first step, and more
work is required to investigate and compare useful
in-band synchronization abstractions, also in terms of performance.
While our framework can be used as a synchronization library, we 
also work on integrating it into an existing SDN controller, e.g., Ryu. 
%Moreover, we will consider adding special
%locking/synchronization primitives to OpenFlow.
